Takes a python or matlab matrix or nested list and converts to a LaTeX table or matrix.
Author: ob@cakebox.net, inspired by the work of koehler@in.tum.de who has written
a similar package only for matlab
\url{http://www.mathworks.com/matlabcentral/fileexchange/4894-matrix2latex}

This software is published under the GNU GPL, by the free software
foundation. See:
\url{http://www.gnu.org/licenses/licenses.html#GPL}

\section{TODO}
\begin{itemize}
\item Improve test suite.
\item Improve error handling.
\item Update matlab code
\end{itemize}

\section{Compatibility}
Table \ref{tab:features} reflect the current features and whether it is available
in the matlab or python version.
\input{features}
These features are discussed in more detail in chapter \ref{sec:python}.

\subsection{Python}
The code is written without any dependencies and should be compatible with most python versions.
Table \ref{tab:compatibleTable} reflects the python versions currently installed
on my system\footnote{Mac OS X 10.10, python installed through macport.}
and if the testsuit for matrix2latex passes. If you find a python version where
it doesn't work let me know.
\input{compatibleTable}
\subsection{Matlab}
For the moment it is only tested with matlab R2009b.

\section{Installation}
The following packages and definitions are recommended in the latex preamble 
\begin{pygments}{latex}
% scientific notation, 1\e{9} will print as 1x10^9
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\usepackage{amsmath} % needed for pmatrix
\usepackage{booktabs} % Fancy tables
\usepackage{caption} % Fixes spacing between caption and table
...
\begin{document}
...
\end{pygments}
\subsection{Python}
To use the code place the folder matrix2latex/ in our \verb!PYTHONPATH!, 
this can be done with the usual
\begin{verbatim}
sudo python setup.py install
\end{verbatim}
Alternativly, your current working directory is always on your \verb!PYTHONPATH!.

% Hint: on unix systems do: \\
% \verb!echo $PYTHONPATH! \\
% to see a list of locations. Other users: ask google about \verb!PYTHONPATH!
% for your operation system.
